BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-04-28 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
format things

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual generation of herbarium specimens is essential for current and future research on ecology, evolution, and conservation. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau for administrative jurisdictional data, and other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions and locality information. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with current taxonomic opinions, generates driving directions to sites, and produces aesthetically pleasing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread]. 
This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.  These factors ensure that the specimens remain the premier botanical data source into perpetuity.

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Evidently apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error. 
Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. Even if a collector navigates all these hurdles, the time allocated to each step is quite large.

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual accessioning of herbarium specimens is essential for current and future research on ecology, evolution, and conservation. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens for collectors.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau, the United States Geological Survey, and other sources to supplement collection notes by providing information on administrative jurisdictions, abiotic site conditions and locality information. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with current taxonomic opinions, generates driving directions to sites, and produces aesthetically pleasing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread]. 
This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences and grearly expanded their roles in science [@ronsted2020integrative; @marsico2020small].

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as electronic and remote sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
These factors will ensure that the specimens remain the premier botanical data source into perpetuity.

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via commercially developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report administrative details in error. 
Even finding appropriate site names can rarely be resolved without a printed map, as many navigation-related software now consider most features that would serve as site names extraneous. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied scientific names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting which reduce the apparent credibility of a collector. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. 
Even if a collector successfully navigates all these hurdles, the time allocated to each step is quite large, and may discourage them from further collecting.
